---
title: "individual1"
author: "Ryan Lode"
date: "9/27/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r,message = FALSE }
library(readxl)
library(tidyverse)
```
```{r, message = FALSE, include = FALSE}
GSSdata <- read_excel('./GSS.xls', sheet='Data')
```

1.Find the average years of education for the respondents with each political party affiliation. Arrange your output in a meaningful order and print. Describe any patterns you find.


First the GSSdata column names are adjusted to include underscores instead
of spaces. Completed education is converted into numeric values and party affiiliation is modified into a factor. Then, party affilation is grouped by and we declare a variable AverageEducation with the mean of the completed education, before displaying it in increasing order. 


```{r, warning=FALSE, message=FALSE}
names(GSSdata) <- c('Gss_year', 'general_happiness', 'Welfare', 'drug_addiction', 'national_health', 'environment','party_affiliation','completed_education','respondent_age','marital_status', 'respondent_id', 'ballot_used')
GSSdata <- GSSdata %>%
  mutate(completed_education = as.numeric(completed_education), 
        party_affiliation = as.factor(party_affiliation))

GSSdata %>%
  group_by(party_affiliation) %>%
  summarise(AverageEducation = mean(completed_education,na.rm = TRUE)) %>%
  arrange(AverageEducation)

```

**Analysis: **From the output, we can we see that those whose political party is Strong Republican have highest mean education level. Those who don't know their party affiliation have the lowest mean education level. 


2.Create a single data frame containing records for the better educated respondents in each political party affiliation. A “better educated” respondent is someone who has strictly more years of education than the median in the same party affiliation. Print the structure of the data frame.

```{r}
betterEducated <- GSSdata %>% group_by(party_affiliation) %>% 
  filter(completed_education > median(completed_education, na.rm = TRUE))

betterEducated <-betterEducated[order(betterEducated$completed_education),]

str(betterEducated)
```

```{r}
head(betterEducated[,4:8])
```

**Analysis : ** Examining the tibble, we can see that the completed education years don't fall below a certain number within political party affiliations.

3.How is the happiness of a respondent related to his/her marriage status? Define that a person is happy if the response to question is “Very happy” or “Pretty happy”.


**Part1: ** A new tibble, HappyVsMarriageStatus is created from GSSdata
Next, the dataframe is cleaned with na.omit and a series
of filter statements.Marital status and party affiliation columns are 
then converted to factors.

```{r}
HappyVsMarriageStatus <-GSSdata

HappyVsMarriageStatus = na.omit(HappyVsMarriageStatus)
HappyVsMarriageStatus <- HappyVsMarriageStatus %>%
  filter(!HappyVsMarriageStatus$general_happiness %in% c('No answer','Not applicable','Don\'t know')) 
HappyVsMarriageStatus <-HappyVsMarriageStatus %>%
  filter(!HappyVsMarriageStatus$marital_status %in% c('No answer'))
HappyVsMarriageStatus<-HappyVsMarriageStatus %>%
  filter(!HappyVsMarriageStatus$party_affiliation %in% c('Don\'t know', 'No answer'))

maritalStatus <-c('Married','Widowed','Divorced','Separated','Never married')
HappyVsMarriageStatus$marital_status <-factor(HappyVsMarriageStatus$marital_status, levels = maritalStatus)

politicalParty <- c('Other party','Strong democrat', 'Not str democrat','Ind,near dem','Independent','Ind,near rep', 'Not str republican','Strong republican')
HappyVsMarriageStatus$party_affiliation <-factor(HappyVsMarriageStatus$party_affiliation, levels = politicalParty)
```
```{r}
HappyVsMarriageStatus %>%
  group_by(general_happiness, marital_status) %>%
  ggplot(aes(x=marital_status, fill= general_happiness)) +
 geom_bar(position = 'fill') +scale_fill_brewer(palette="Paired") + ggtitle('Marital Status vs. Happiness') + 
 ylab('Frequency') + xlab('Marital Status')
```

**Part2: **A new tibble GSSdataVeryHappy was created by
calling count on columns marital_status and general_happiness
and then grouping by those two columns. A new column indicating the
denoting the frequency of each occurrance was generated as well.

```{r}
GSSdataVeryHappy <-HappyVsMarriageStatus %>%
  count(marital_status, general_happiness) %>%
  group_by(marital_status, general_happiness) %>%
  ungroup

GSSdataVeryHappy <-GSSdataVeryHappy  %>%
  mutate((freq = n /sum(GSSdataVeryHappy$n)) * 100)


GSSdataVeryHappy <-GSSdataVeryHappy %>%
  filter(GSSdataVeryHappy$general_happiness %in% c('Very happy', 'Pretty happy'))

names(GSSdataVeryHappy)[names(GSSdataVeryHappy) == "(freq = n/sum(GSSdataVeryHappy$n)) * 100"] <- 'Relative Frequency'
GSSdataVeryHappy <-GSSdataVeryHappy[order(GSSdataVeryHappy$`Relative Frequency`, decreasing = TRUE), ]
```

```{r}
head(GSSdataVeryHappy)

```

**Analysis: **Analysis: For the general population, of those who reported being either pretty happy' or 'very happy', the majority of them were married. Respondents who were separated from their partners appear to have the lowest percentage of people, among the different categories of marital status, to report being 'Very Happy'.

4.Does party affiliation affect the relationship you found in the last question?

General Happiness and marital status were grouped by marital status and
general happiness and then separated by political party. 
The resulting plot shows the  proportions of people, given a political
party and marriage status, who are either very happy, pretty happy,
or not too happy.A new tibble 'Strong Republicans' was created that 
focuses solely on people with that particular party affiliation.
StrongRepublicans is further filtered to only include married
Republicans who listed themselves as either 'Very happy' or 'Pretty happy'A relative frequency column was added as well. 

```{r,fig.width=10, fig.height=4}
HappyVsMarriageStatus %>%
  group_by(general_happiness, marital_status) %>%
  ggplot(aes(x=marital_status, fill= general_happiness)) + geom_bar(position = 'fill') +scale_fill_brewer(palette="Paired") + theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  facet_grid(~party_affiliation) +ylab('Frequency') + xlab('Marital Status') +
  ggtitle('Marital Status vs. Happiness by Party Affiliation')
```

```{r}
StrongRepublicans <- HappyVsMarriageStatus %>%
  count(marital_status, general_happiness, party_affiliation) %>%
  group_by(marital_status,general_happiness, party_affiliation) %>%
  ungroup
StrongRepublicans <-StrongRepublicans %>%
  filter(StrongRepublicans$party_affiliation %in% c('Strong republican'))
MarriedRepublicans <-StrongRepublicans %>%
  filter(StrongRepublicans$marital_status %in% c('Married'))

MarriedRepublicans <-MarriedRepublicans %>%
  mutate((freq = n /sum(MarriedRepublicans$n)) * 100) %>% 
  ungroup
  
names(MarriedRepublicans)[names(MarriedRepublicans) == "(freq = n/sum(MarriedRepublicans$n)) * 100"] <- 'Relative Frequency %'

MarriedRepublicans <-MarriedRepublicans[order(MarriedRepublicans$`Relative Frequency %`, decreasing = TRUE),]


```

```{r}
print(MarriedRepublicans[,])
```

**Analysis: **Examining the frequency plot, Strong republicans seem to have an edge in terms of overall happiness. Filtering by Strong republicans and married shows that married republicans are overall, quite happy. In fact, over 95% of married Republicans report being either 'Very happy' 'pretty happy'.


5.A histogram and boxplot were generated from the tibble HappyVsMarriageStatus, both of which show the distribution of the education level in years for all survey respondents.Then summary education statistics were generated for the 'Divorced'and 'Widowed' categories. 

```{r}
HappyVsMarriageStatus %>%
  group_by(marital_status) %>%
  ggplot(aes(x=completed_education)) +geom_histogram() + facet_grid(~marital_status) +
  ylab('Number of Respondents') + xlab('Completed Education(Years)') + ggtitle('Distribution of Completed Education by Marital Status')
```


```{r}
HappyVsMarriageStatus %>%
  group_by(marital_status) %>%
  ggplot(aes(x=completed_education)) +geom_boxplot() + facet_grid(~marital_status,)+ coord_flip()+
  xlab('Completed Education (Years)') + theme(axis.text.x = element_blank(),axis.ticks.x  = element_blank()) + 
  ggtitle('Boxplot Distribution of Completed Education by Marital Status')

```

```{r}
divorcedEducation<-HappyVsMarriageStatus %>%
  filter(marital_status %in% c('Divorced'))

NeverMarriedEduc <-HappyVsMarriageStatus %>%
  filter(marital_status %in% c('Never married'))

marriedEducation <-HappyVsMarriageStatus %>%
  filter(marital_status %in% c('Married'))
```

```{r}
summary(divorcedEducation$completed_education)
summary(NeverMarriedEduc$completed_education)
summary(marriedEducation$completed_education)
```

**Analysis: **Analysis:Examining the boxplots and summary statistics, it appears as though divorced respondents had the highest median education level. Married people had a median education level identical to the Q1 value. This is probably due to the very large spike in 
seen in the histogram. Those who were never married report the highest(tied with those) who were divorced) median education level and the highest Q3 value. Perhaps those who chose not to marry instead focused on continuing their education or research.


6a.How does political party affiliation impact views on the environment? What is the education level distribution of those who think the government is doing too little to protect the environment?

A new tibble environment was created.Then, the environmental column was 
factorized. Several filter statements were then used to clean the data.
A plot showing the proportions environmental views was then generated with ggplot. Finally, a new tibble TooLittle was created that only contained data from those who answered whether the government was doing 'Too little' to protect the environment. Summary statistics were then generated on the education level of the respondents.

```{r}
environment <-GSSdata


politicalParty2 <- c('Other party','Strong democrat', 'Not str democrat','Ind,near dem','Independent','Ind,near rep', 'Not str republican','Strong republican')
environment$party_affiliation <-factor(environment$party_affiliation, levels = politicalParty2)

environmentalCare <- c('Too much','About right','Too little')
environment$environment<-factor(environment$environment, levels = environmentalCare)
environment <-environment %>%
  filter(!is.na(environment))

environment <-environment %>%
  filter(!environment %in% c('No answer','Don\'t know','Not applicable', 'NA'))

environment <-environment %>%
  filter(!party_affiliation %in% c('Don\'t know','No answer', 'NA'))

environment = na.omit(environment)

environment <-environment %>%
  filter(!party_affiliation %in% c('Don\'t know','No answer', 'NA'))

environment %>%
  group_by(party_affiliation) %>%
  ggplot(aes(x = party_affiliation, fill = environment)) + geom_bar(position = 'fill') + theme(axis.text.x = element_text(angle = 90, hjust = 1)) +ggtitle('Party Affiliation vs Environmental Views')
```

```{r}
TooLitte <-environment
```

```{r}
TooLitte <-TooLitte %>%
  filter(environment %in% c('Too little'))
summary(TooLitte$completed_education)
```

**Analysis: ** Examining the plot, the proportions of respondents who
think the United States is doing too little to protect the environment
declines as the views become more conservative. This makes sense given
the party's historical platform. Summary statistics of those who think
the government is doing too little to protect the environment, show an 
IQR of 3 years, but the median was the same as Q1-12 years. 

6b.)How does political party affiliation impact views on welfare? What is the mean education level given that the respondents view
the government supports welfare too much?

A new tibble, Welfare, was created, then it's welfare and political party columns were factorized. A series of filter statemnts were then used to clean the data. A staggered bar chart showing the counts of Welfare views
by political party was then generated.Finally, a new tibble called TooMuch was generated that focused on those respondents who view that the government is supporting welfare too much. Given that 'Strong Republicans' had the highest mean education level, we might expect the mean education level in the TooMuch tibble might be high. 

```{r}
Welfare <-GSSdata



Welfare <-Welfare %>%
  mutate(Welfare = as.factor(Welfare))



GeneralWelfare <- c('Too little','About right','Too much')

Welfare$Welfare<-factor(Welfare$Welfare, levels = GeneralWelfare)

politicalParty3 <- c('Other party','Strong democrat', 'Not str democrat','Ind,near dem','Independent','Ind,near rep', 'Not str republican','Strong republican')
Welfare$party_affiliation <-factor(Welfare$party_affiliation, levels = politicalParty3)

Welfare <-Welfare %>%
  filter(!Welfare %in% c('Don\'t know','No answer', 'NA', 'Not applicable'))

Welfare <-Welfare %>%
  filter(!party_affiliation %in% c('Don\'t know','No answer', 'NA'))

Welfare <-Welfare %>%
  filter(!is.na(Welfare))

Welfare = na.omit(Welfare)

Welfare %>%
  group_by(party_affiliation) %>%
  ggplot(aes(x = party_affiliation, fill = Welfare)) + geom_bar(position = 'dodge') +ggtitle('Party Affiliation vs Welfare Support')+xlab('Political Party') + theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

```{r}
TooMuch <-Welfare

TooMuch <-TooMuch %>%
  filter(Welfare %in% c('Too much'))


TooMuch <-TooMuch %>%
  mutate(completed_education = as.numeric(completed_education))

TooMuch <-TooMuch %>%
  filter(!is.na(TooMuch))

summary(TooMuch$completed_education)

```

**Analysis: ** Examining the bar chart, relative to the other political parties, those in the Strong republican and Not str republican categories think the government supports welfare too much. Interestingly, a distinct majority of Not str democrats think that the government is doing too much to help the poor! The mean education level for those who think that Welfare support is too high is 12.82.The median, however, is a 12-indicating they only completed high school. 
